---
title: "Churn Analysis"
author: "Nicholas Bibeau"
date: "March 18, 2019"
output: 
  html_document:
    toc: TRUE
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(polycor)
library(corrplot)
```

# Data Collection
```{r}
churn <- read_csv("https://www.iun.edu/~cisjw/ds/files/data/churn_customer.csv", col_types = cols(
  Technology = col_character(),
  Age = col_double(),
  CustomerSince = col_character(),
  #CustomerSince = col_datetime(format='%m, %d, %y'),
  SupportCallsLastYear = col_double(),
  AverageBill = col_double(),
  ChurnIndicator = col_double()
))
summary(churn)
```

# Churn Indicator

```{r}
#Add discrete labels in a new column
churn <- churn %>% mutate(churn_label=ifelse(ChurnIndicator>0.5,1,0))
#histogram: frequency plot of ChurnIndicator colored by churn label
churn %>%
  ggplot(aes(x=ChurnIndicator,fill=as.factor(churn_label)))+
  geom_bar()+
  scale_y_log10()+
  labs(y="log10(count)")

```

## Write the dataset to a file

```{r}
write_csv(churn, file.path("churn_v2.csv"), quote_escape = "double")
```

## Find distribution of churn and no churn
```{r}
churn %>%
  group_by(churn_label) %>%
  summarise(n())

```

# Correlation Analysis

## Generating Correlation Matrix
```{r}
churn.cor <- hetcor(churn$ChurnIndicator, churn$churn_label, churn$Technology, churn$Age, churn$AverageBill, churn$SupportCallsLastYear)
churn.cor$correlations
```

## Visualizing the correlation matrix in a correlogram
```{r}
corrplot(churn.cor$correlations, method="shade", type="upper", tl.col="black", tl.srt=45) 
```

## Producing a rank of the regular attributes in terms of the correlation strength with churnIndicator in descending order
```{r}
names <- c("technology", "age", "callsLastYear", "averageBill", "churnIndicator", "churn_label")
churn.cormat <- as_tibble(churn.cor$correlations)
colnames(churn.cormat) <- names

churn.cormat <- churn.cormat %>% 
  mutate(attribute=names) %>% 
  select(attribute,churnIndicator,callsLastYear,churn_label,averageBill,technology,age) 

rank <- churn.cormat %>% 
  select(attribute, churnIndicator) %>% 
  mutate(squared.correlation=churnIndicator^2) %>%
  arrange(desc(squared.correlation)) 

rank
```

# Normalizing Data

## Create Normalization Functions
```{r}
#creating normalization function
normalize <- function(v, range=1){
    v.norm = (v - min(na.omit(v))) / (max(na.omit(v))-min(na.omit(v))) * range
    return(v.norm)
}

normalizeDataset <- function (data){
  data.norm <- data
  types <- sapply(data, is.numeric)
  for (i in 1:length(types)) {
    if(types[i]==TRUE){
      data.norm[ ,i] <- normalize(data[ ,i])
    }
  }
  return(data.norm)
}
```

## Data Normalization
```{r}
# Normalizing train dataset
churn.norm <- normalizeDataset(churn)
churn.norm
```

# Analysis
```{r}
churn %>%
  group_by(Technology) %>%
  summarise(n())
```

# Readying Data for Regression Analysis

## Changing Attribute Names to Lowercase
```{r}
colnames(churn.norm) <- tolower(colnames(churn.norm))
summary(churn.norm)
```

## Split the set into two sets for the training and testing purposes
```{r}
set.seed(52)
rows <- sample(nrow(churn.norm)) #Shuffle row indices
rows

churn.norm <- churn.norm[rows, ] #Randomly reorder the churn.norm rows by the vector rows
churn.norm
```

## Subset data
```{r}
split <- round(nrow(churn.norm)*0.66)

# Put first 66% into train dataset
train <- churn.norm[1:split, ]

# Put remainining 1/3 into test dataset
test <- churn.norm[(split+1):nrow(churn.norm), ]
```

# Fit model
```{r}
model <- lm(churnindicator ~ ., train) 

summary(model)
```

## Extracting coefficients and r-squared
```{r}
#Extract Coefficients
model$coefficients #Or coef(model)

#Read r-squared
summary(model)$r.squared
```

# Plot fitted values vs. true values in the training set
```{r}
plot(model$fitted.values, train$churnindicator, xlab="true_ChurnIndicator", ylab="fitted_ChurnIndicator", main="t")
```

#  Calculate RMSE in the training set
```{r}
# Calculating the errors between every true value and its fitted value in the training set and store the errors in a vector error:
error <- model$fitted.values - train$churnindicator

# Calculating RMSE:
rmse.train <- sqrt(mean(error^2))
rmse.train
```

# Applying the model to the test set
```{r}
# We can apply the model to the test set test by the predict function. Store the predictions in a vector called scores:
scores <- predict(model, test)
scores
```

# Calculating Error Metric: RMSE (root mean square error)
```{r}
# Calculating the errors between every true value and its fitted value in the training set and store the errors in a vector error:
error <- scores - test$churnindicator
# Calculating RMSE:
rmse.test <- sqrt(mean(error^2))
rmse.test
```

# Comparing rmse.train and rmse.test
```{r}
# TBD
```



