---
title: "Exam 1"
author: "Nicholas Bibeau"
date: "February 25, 2019"
output: 
  html_document:
    toc: TRUE
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

#creating normalization function
normalize <- function(v, range=1){
    v.norm = (v - min(na.omit(v))) / (max(na.omit(v))-min(na.omit(v))) * range
    return(v.norm)
}

normalizeDataset <- function (data){
  data.norm <- data
  types <- sapply(data, is.double)
  for (i in 1:length(types)) {
    if(types[i]==TRUE){
      data.norm[ ,i] <- normalize(data[ ,i])
    }
  }
  return(data.norm)
}
```

# 1. Import Dataset
```{r}
sales <- read_csv("https://www.iun.edu/~cisjw/ds/files/data/yearly_sales.csv")
```

# 2. Write code to return the total rows and total columns
```{r}
# Number of rows
nrow(sales)

# Number of columns
ncol(sales)

# Rows and columns as dimensions
dim(sales)
```

# 3. Write a single statement to find the data type of each column.
```{r}
#sales %>% summarize(class(c(1:4, ))
#c(class(sales$cust_id, sales$sales_total, sales$num_of_orders, sales$gender))
class(sales$cust_id)
class(sales$sales_total)
class(sales$num_of_orders)
class(sales$gender)

```

# 4. According to your understanding of the column values, be sure that the tibble has every column set in the right data type
```{r}
sales <- read_csv("https://www.iun.edu/~cisjw/ds/files/data/yearly_sales.csv", col_types = cols(
  cust_id = col_integer(),
  sales_total = col_double(),
  num_of_orders = col_double(),
  gender = col_character()
))
```

# 5. Write a statement which returns a boolean vector, with the ith element being either TRUE or FALSE
```{r}

```

# 6. Display the first five rows in sales
```{r}
head(sales, 5)
```

# 7. Find the total males and females, respectively.
```{r}
# Total males
sales %>% filter(gender == "M") %>% count()

# Total females
sales %>% filter(gender == "F") %>% count()
```

# 8. For both male and female groups, find their average, maximum, and variance on sales_total and num_of_orders, respectively. Give each value a name.
```{r}
m_sales_total_avg <- sales %>% filter(gender == "M") %>% summarize(mean(sales_total))
m_sales_total_max <- sales %>% filter(gender == "M") %>% summarize(max(sales_total))
m_sales_total_var <- sales %>% filter(gender == "M") %>% summarize(var(sales_total))
m_num_of_orders_avg <- sales %>% filter(gender == "M") %>% summarize(mean(num_of_orders))
m_num_of_orders_max <- sales %>% filter(gender == "M") %>% summarize(max(num_of_orders))
m_num_of_orders_var <- sales %>% filter(gender == "M") %>% summarize(var(num_of_orders))
f_sales_total_avg <- sales %>% filter(gender == "F") %>% summarize(mean(sales_total))
f_sales_total_max <- sales %>% filter(gender == "F") %>% summarize(max(sales_total))
f_sales_total_var <- sales %>% filter(gender == "F") %>% summarize(var(sales_total))
f_num_of_orders_avg <- sales %>% filter(gender == "F") %>% summarize(mean(num_of_orders))
f_num_of_orders_max <- sales %>% filter(gender == "F") %>% summarize(max(num_of_orders))
f_num_of_orders_var <- sales %>% filter(gender == "F") %>% summarize(var(num_of_orders))
```

# 9. For each gender group, find the total of rows who have sales_total greater than 3K.
```{r}
# Males with sales totals greater than 3k
sales %>% filter(gender == "M", sales_total > 3000) %>% count()

# Females with sales totals greater than 3k
sales %>% filter(gender == "F", sales_total > 3000) %>% count()
```

# 10. For each gender group, find the total of rows who have sales_total greater than 3K and num_of_orders less than 10.
```{r}
# Males with sales totals greater than 3k and fewer than 10 orders
sales %>% filter(gender == "M", sales_total > 3000, num_of_orders < 10) %>% count()

# Females with sales totals greater than 3k and fewer than 10 orders
sales %>% filter(gender == "F", sales_total > 3000, num_of_orders < 10) %>% count()
```

# 11. Perform normalization to the columns in sales, which you believe should be normalized to the same range between zero and unit one.
```{r}
sales.norm <- normalizeDataset(sales)

```

