---
title: "Exercise 5 - Bibeau"
author: "Nicholas Bibeau"
date: "January 31, 2019"
output: 
  html_document:
    toc: TRUE
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# 5 Exercise

## 5.1 E1
```{r 5.1E1}
#Write R code to create the following data frame of 5 rows and 5 named columns, store the data in a name weather, and print the data.
weather <- data.frame( 
 Outlook = c("sunny", "sunny", "overcast", NULL, NULL),
 Temperature = c(85, 80, 83, NULL, NULL),
 Humidity = c(85, 90, 86, NULL, NULL),
 Windy = c(FALSE, TRUE, FALSE, NULL, NULL),
 Play = c("no", "no", "yes", NULL, NULL)
 )

print(weather)
```


## 5.2 E2
```{r 5.2E2}
# Import built-in iris dataset
iris
```

### 5.2.1 Opening the help file of a built-in dataset
```{r 5.2.1}
help(iris)
```

### 5.2.2 Examining the object type
```{r 5.2.2}
class(iris)
```

### 5.2.3 Retrieving a column from a dataframe

#### Retrieving Species column from iris dataset
```{r}
iris$Species
```

#### Retrieving Petal.Length column from iris dataset
```{r}
iris$Petal.Length
```

### 5.2.4 Finding data type of a column (attribute)

#### Finding the type of column vector Species in iris
```{r irisColumnType1}
class(iris$Species)
```

#### Finding the type of column vector Sepal.Length in iris
```{r irisColumnType2}
class(iris$Sepal.Length)
```

### 5.2.5 Subscripting

#### Retrieving first 50 rows of iris dataset
```{r 5.2.5}
iris[1:50,]
```


#### Retrieve the first 2 columns (attributes) of iris
```{r irisColumns1and2}
iris[, 1:2]
```

#### Retrieving the columns Sepal.Length and Petal.Length
```{r irisPetalLenthAndSepalLength}
iris[, c("Sepal.Length", "Petal.Length")]
```

#### Retrieving All of the columns excluding the last column Species
```{r irisExceptSpecies}
iris[, colnames(iris) !="Species"]
```

#### Viewing rows for Species 'setosa'
```{r irisSetosa}
iris[iris$Species == "setosa", ]
```


### 5.2.6 Displaying column-wise statistics data
```{r 5.2.6}
summary(iris)

# Min. represents the smallest value in each column.
# 1st Qu. represents the maximum value in the lowest 25% of values in each column (the upper limit of the 1st quartile.)
# Median represents the midpoint in the distribution of values in each column (the upper limit of the 2nd quartile.)
# Mean represents the average of values in each column.
# 3rd Qu. represents the maximum value in the 50-75% range of values in each column (the upper limit of the 3rd quartile.)
# Max represents the largest value in each column.
```

## 5.3 E3
```{r 5.3E3}
df <- data.frame(X = -2:2, Y = 1:5)
result1 <- df$Y[ df$X > 0 ]
result2 <- df[ df$X > 0, ]

# result1 will provide a vector of Y-values from df corresponding to X-values greater than 0.
# result2 will output a data.frame of X-values in df where X > 0, along with corresponding Y-values.
```

## 5.4 E4 mtcars

### Importing the built-in mtcars dataset
```{r 5.4E4}
mtcars
```

### Understanding mtcars dataset by using help
```{r mtcarsHelp}
help(mtcars)
```

### First five rows of mtcars dataset
```{r mtcarsHead}
head(mtcars, 5)
```

### Last five rows of mtcars dataset
```{r mtcarsTail}
tail(mtcars, 5)
```

### Dimensions of mtcars dataset
```{r mtcarsDimensions}
dim(mtcars)
# mtcars has 32 rows and 11 columns
```

### Viewing mtcars dataset
```{r mtcarsView}
data(mtcars)
View(mtcars)
```

### MPG
```{r mtcarsMPG}
mtcars$mpg
```

### MPG of 6-cylinder vehicles
```{r mtcars6CylMPG}
mtcars$mpg[mtcars$cyl == 6]
```

### 6-cylinder vehicles
```{r mtcars6Cyl}
mtcars[mtcars$cyl == 6, ]
```


### Printing rows of where mpg is greater than 25, but only for the mpg and cyl columns.
```{r mtcarsMPG>25}
mtcars[mtcars$mpg > 25, c("mpg", "cyl")]
```

## 5.5 E5
```{r 5.5E5}
# Print the built-in diamonds data
diamonds
```

### Viewing help for diamonds dataset
```{r diamondsHelp}
help(diamonds)
```

### Viewing first five rows of the diamonds dataset
```{r diamondsHead}
head(diamonds, 5)
```

### Finding dimensions (rows and columns) of the diamonds dataset
```{r diamondsDimensions}
dim(diamonds)
# There are 53840 rows and 10 columns in diamonds
```

### Diamonds of cut "Very Good"
```{r diamondsCutVeryGood}
diamonds[diamonds$cut == "Very Good", ]
# There are 12,082 rows where cut equals Very Good.
```

### Diamonds larger than 3.0 carat
```{r diamondsCaratGreaterThan3}
count(diamonds[diamonds$carat > 3.0, ])
# There are 32 diamonds with carat greater than 3.0
```

### Diamonds of color "D"
```{r diamondsColorD}
diamonds[diamonds$color == "D", c("color", "cut"), drop=FALSE]
```

### Diamonds Summary
```{r diamondsSummary}
summary(diamonds)
# The average diamond price is $3933.
```

