---
title: "INFO-I 421 Final - Building Decision Tree Models for the Breast Cancer Data"
author: "Nicholas Bibeau"
date: "April 29, 2019"
output: 
  html_document:
    toc: TRUE
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(polycor)
library(corrplot)
library(rpart)
library(rpart.plot)
library(DiagrammeR)
library(useful)
library(xgboost)
library(vip)
library(randomForest)

# User-defined function to 
# 1. calculate Shannon entropy for a given vector 
# 2. visualize the distribution in a histogram

entropy.gini <- function(input, type=1, discrete=1){
  # a function to calculate Shannon entropy for a given vector
  # type: 1- categories in integers, continuous values
  # type: 2- class counts
  # when type=1, discrete: 1-discrete labels, 0-continuous labels
  
  if (type==1){
    if (discrete == 0){
      breaks <- c(floor(min(input)):ceiling(max(input)))
    } else {
      breaks <- c(min(input):(max(input)+1))
    }
    freq <- hist(input,include.lowest=TRUE,right=FALSE, breaks=breaks,col='lightblue')
    counts <- freq$counts
    total <- sum(counts)
    
  }
  else{
    breaks <- length(input)
    total <- sum(input)
    counts <- input
    hist(input)
  }
  
  entropy <- 0
  for(x in counts){
    if(x!=0)
      temp <- x/total*log2(x/total)
    else
      temp <- 0

    entropy <- entropy - temp
  }
  
  gini <- 1-sum((counts/total)**2)
  
  mylist <- list("breaks"=breaks,"entropy"=entropy,"gini"=gini)
  return(mylist)
}
```

# 1.1 Importing Data
```{r data_import}
breast_cancer <- as_tibble(read.table("https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer/breast-cancer.data", header = FALSE, na = "?", sep = ",", col.names=c("class", "age", "menopause", "tumor_size", "inv_nodes", "node_caps", "deg_malig", "breast", "breast_quad", "irradiat")))
```

# 1.2 Finding the age category which has the highest risk of recurrence

## Viewing distinct factors for age
```{r unique_age}
unique(breast_cancer$age)
```

## Building contingency table for age and recurrence
```{r age_recur_table}
breast_cancer_table <- table(breast_cancer$age, breast_cancer$class, dnn = c("Age", "Recurrence"))
breast_cancer_table
```

## Calculating probability of recurrence by age group

```{r prob_recur_by_age}
probAge1 <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(age == "20-29")) /
  count(breast_cancer %>% filter(age == "20-29"))
probAge2 <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(age == "30-39")) /
  count(breast_cancer %>% filter(age == "30-39"))
probAge3 <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(age == "40-49")) /
  count(breast_cancer %>% filter(age == "40-49"))
probAge4 <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(age == "50-59")) /
  count(breast_cancer %>% filter(age == "50-59"))
probAge5 <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(age == "60-69")) /
  count(breast_cancer %>% filter(age == "60-69"))
probAge6 <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(age == "70-79")) /
  count(breast_cancer %>% filter(age == "70-79"))

# Creating vector of recurrence probability by age group
recur_probs <- c(probAge1, probAge2, probAge3, probAge4, probAge5, probAge6)

# Creating vector of distinct age groups
ages <- c ("20-29", "30-39", "40-49", "50-59", "60-69", "70-79")

# Creating named vector of recurrence probability by age group
names(recur_probs) <- ages

```

## Determining age category with the highest risk of recurrence
```{r max_risk_age} 
# The age group with the highest risk of recurrence is: 
names(recur_probs)[which.max(recur_probs)]
```

# 1.3 Finding breast with the higher risk of recurrence

## Viewing distinct factors for breast
```{r unique_breast}
unique(breast_cancer$breast)
```

## Building contingency table for age and recurrence
```{r breast_recur_table}
breast_cancer_table <- table(breast_cancer$breast, breast_cancer$class, dnn = c("Breast", "Recurrence"))
breast_cancer_table
```

## Calculating probability of recurrence by breast

```{r prob_recur_by_breast}
probLeft <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(breast == "left")) /
  count(breast_cancer %>% filter(breast == "left"))
probRight <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(breast == "right")) /
  count(breast_cancer %>% filter(breast == "right"))

# Creating vector of recurrence probability by breast
recur_probs_breasts <- c(probLeft, probRight)

# Creating vector of distinct breasts
breasts <- c ("left", "right")

# Creating named vector of recurrence probability by age group
names(recur_probs_breasts) <- breasts

```

## Determining breast with the higher risk of recurrence
```{r max_risk_breast} 
# The age group with the highest risk of recurrence is: 
names(recur_probs_breasts)[which.max(recur_probs_breasts)]
```

# 1.4 Drop missing values
```{r drop_missing}
breast_cancer <- breast_cancer %>% drop_na()
```

# 1.5 Rank attributes and plot correlation matrix

## Building correlation matrix
```{r correlation_matrix}
# All vectors in breast_cancer dataset are numeric or factors, so no mutations are necessary.
breast_cancer.cor <- hetcor(breast_cancer$class, breast_cancer$age, breast_cancer$menopause, breast_cancer$tumor_size, breast_cancer$inv_nodes, breast_cancer$node_caps, breast_cancer$deg_malig, breast_cancer$breast, breast_cancer$breast_quad, breast_cancer$irradiat)

```

## Storing correlation matrix with updated column names
```{r update_matrix}
# Creating correlation matrix
breast_cancer.corrmat <- as_tibble(breast_cancer.cor$correlations)

# Assigning original column names to correlation matrix
colnames(breast_cancer.corrmat) <- colnames(breast_cancer)

# Viewing correlation matrix
breast_cancer.corrmat
```

## Adding attribute names in a new column and reordering columns
```{r tweaking_columns}
breast_cancer.corrmat <- breast_cancer.corrmat %>% 
  mutate(attribute=colnames(breast_cancer.corrmat)) %>% 
  select(attribute, class, age, menopause, tumor_size, inv_nodes, node_caps, deg_malig, breast, breast_quad, irradiat)
```

## Printing the ranking
```{r rank_order}
rank <- breast_cancer.corrmat %>% 
  select(attribute, class) %>% 
  mutate(squared.correlation=class^2) %>%
  arrange(desc(squared.correlation)) 

rank
```

## Visualizing the correlation matrix in a correlogram

```{r correlogram}
corrplot(breast_cancer.cor$correlations, method="shade", type="upper", tl.col="black", tl.srt=45)
```

# 1.6 Decision tree models

## 1.6.1 Calculation information gain for each regular attribute

### Computing entropy of dataset for gain calculations
```{r total_entropy}
#Finding counts for entropy calculation
counts <- (breast_cancer %>% 
  select(class) %>%
  group_by(class) %>%
  summarise(count=n())
  )$count

# Calculating entropy and gini
entropy_set <-entropy.gini(counts)

# The entropy for the entire dataset is: 
entropy_set$entropy
```

### Calculating gain based on age

#### Looking at distribution in terms of three subsets by age
```{r subsetting_age}
info.before <- entropy_set$entropy
subsets.age <- table(breast_cancer$age, breast_cancer$class)
subsets.age
```

#### Computing entropy of age subsets
```{r entropy_age}
entropy20_29 <-entropy.gini(subsets.age["20-29",], type=2)$entropy
entropy30_39 <-entropy.gini(subsets.age["30-39",], type=2)$entropy
entropy40_49 <-entropy.gini(subsets.age["40-49",], type=2)$entropy
entropy50_59 <-entropy.gini(subsets.age["50-59",], type=2)$entropy
entropy60_69 <-entropy.gini(subsets.age["60-69",], type=2)$entropy
entropy70_79 <-entropy.gini(subsets.age["70-79",], type=2)$entropy
```

#### Computing weighted entropy for age
```{r weighted_entropy_age}
total <- nrow(breast_cancer)
w20_29 <- sum(subsets.age["20-29",])/total
w30_39 <- sum(subsets.age["30-39",])/total
w40_49 <- sum(subsets.age["40-49",])/total
w50_59 <- sum(subsets.age["50-59",])/total
w60_69 <- sum(subsets.age["60-69",])/total
w70_79 <- sum(subsets.age["70-79",])/total
info.age <- w20_29*entropy20_29 + w30_39*entropy30_39 + w40_49*entropy40_49 + w50_59*entropy50_59 + w60_69*entropy60_69 + w70_79*entropy70_79
```

#### Computing information gain for age
```{r information_gain_age}
gain.age <- info.before - info.age
gain.age
```

### Calculating gain based on menopause

#### Looking at distribution in terms of three subsets by menopause
```{r subsetting_menopause}
subsets.menopause <- table(breast_cancer$menopause, breast_cancer$class)
subsets.menopause
```

#### Computing entropy of menopause subsets
```{r entropy_menopause}
entropy_premeno <-entropy.gini(subsets.menopause["premeno",], type=2)$entropy
entropy_ge40 <-entropy.gini(subsets.menopause["ge40",], type=2)$entropy
entropy_lt40 <-entropy.gini(subsets.menopause["lt40",], type=2)$entropy

```

#### Computing weighted entropy for menopause
```{r weighted_entropy_menopause}
w_premeno <- sum(subsets.menopause["premeno",])/total
w_ge40 <- sum(subsets.menopause["ge40",])/total
w_lt40 <- sum(subsets.menopause["lt40",])/total
info.menopause <- w_premeno*entropy_premeno + w_ge40*entropy_ge40 + w_lt40*entropy_lt40
```

#### Computing information gain for menopause
```{r information_gain_menopause}
gain.menopause <- info.before - info.menopause
gain.menopause
```

### Calculating gain based on tumor size

#### Looking at distribution in terms of three subsets by tumor size
```{r subsetting_tumor_size}
subsets.tumor_size <- table(breast_cancer$tumor_size, breast_cancer$class)
subsets.tumor_size
```

#### Computing entropy of tumor size subsets
```{r entropy_tumor_size}
entropy_ts01 <-entropy.gini(subsets.tumor_size["0-4",], type=2)$entropy
entropy_ts02 <-entropy.gini(subsets.tumor_size["5-9",], type=2)$entropy
entropy_ts03 <-entropy.gini(subsets.tumor_size["10-14",], type=2)$entropy
entropy_ts04 <-entropy.gini(subsets.tumor_size["15-19",], type=2)$entropy
entropy_ts05 <-entropy.gini(subsets.tumor_size["20-24",], type=2)$entropy
entropy_ts06 <-entropy.gini(subsets.tumor_size["25-29",], type=2)$entropy
entropy_ts07 <-entropy.gini(subsets.tumor_size["30-34",], type=2)$entropy
entropy_ts08 <-entropy.gini(subsets.tumor_size["35-39",], type=2)$entropy
entropy_ts09 <-entropy.gini(subsets.tumor_size["40-44",], type=2)$entropy
entropy_ts10 <-entropy.gini(subsets.tumor_size["45-49",], type=2)$entropy
entropy_ts11 <-entropy.gini(subsets.tumor_size["50-54",], type=2)$entropy
```

#### Computing weighted entropy for tumor_size
```{r weighted_entropy_tumor_size}
w_ts01 <- sum(subsets.tumor_size["0-4",])/total
w_ts02 <- sum(subsets.tumor_size["5-9",])/total
w_ts03 <- sum(subsets.tumor_size["10-14",])/total
w_ts04 <- sum(subsets.tumor_size["15-19",])/total
w_ts05 <- sum(subsets.tumor_size["20-24",])/total
w_ts06 <- sum(subsets.tumor_size["25-29",])/total
w_ts07 <- sum(subsets.tumor_size["30-34",])/total
w_ts08 <- sum(subsets.tumor_size["35-39",])/total
w_ts09 <- sum(subsets.tumor_size["40-44",])/total
w_ts10 <- sum(subsets.tumor_size["45-49",])/total
w_ts11 <- sum(subsets.tumor_size["50-54",])/total
info.tumor_size <- w_ts01*entropy_ts01 + w_ts02*entropy_ts02 + w_ts03*entropy_ts03 + w_ts04*entropy_ts04 + w_ts05*entropy_ts05 + w_ts06*entropy_ts06 + w_ts07*entropy_ts07 + w_ts08*entropy_ts08 + w_ts09*entropy_ts09 + w_ts10*entropy_ts10 + w_ts11*entropy_ts11
```

#### Computing information gain for tumor_size
```{r information_gain_tumor_size}
gain.tumor_size <- info.before - info.tumor_size
gain.tumor_size
```

### Calculating gain based on inv_nodes

#### Looking at distribution in terms of three subsets by inv_nodes
```{r subsetting_inv_nodes}
subsets.inv_nodes <- table(breast_cancer$inv_nodes, breast_cancer$class)
subsets.inv_nodes
```

#### Computing entropy of inv_nodes subsein
```{r entropy_inv_nodes}
entropy_in01 <-entropy.gini(subsets.inv_nodes["0-2",], type=2)$entropy
entropy_in02 <-entropy.gini(subsets.inv_nodes["3-5",], type=2)$entropy
entropy_in03 <-entropy.gini(subsets.inv_nodes["6-8",], type=2)$entropy
entropy_in04 <-entropy.gini(subsets.inv_nodes["9-11",], type=2)$entropy
entropy_in05 <-entropy.gini(subsets.inv_nodes["12-14",], type=2)$entropy
entropy_in06 <-entropy.gini(subsets.inv_nodes["15-17",], type=2)$entropy
entropy_in07 <-entropy.gini(subsets.inv_nodes["24-26",], type=2)$entropy
```

#### Computing weighted entropy for inv_nodes
```{r weighted_entropy_inv_nodes}
w_in01 <- sum(subsets.inv_nodes["0-2",])/total
w_in02 <- sum(subsets.inv_nodes["3-5",])/total
w_in03 <- sum(subsets.inv_nodes["6-8",])/total
w_in04 <- sum(subsets.inv_nodes["9-11",])/total
w_in05 <- sum(subsets.inv_nodes["12-14",])/total
w_in06 <- sum(subsets.inv_nodes["15-17",])/total
w_in07 <- sum(subsets.inv_nodes["24-26",])/total
info.inv_nodes <- w_in01*entropy_in01 + w_in02*entropy_in02 + w_in03*entropy_in03 + w_in04*entropy_in04 + w_in05*entropy_in05 + w_in06*entropy_in06 + w_in07*entropy_in07
```

#### Computing information gain for inv_nodes
```{r information_gain_inv_nodes}
gain.inv_nodes <- info.before - info.inv_nodes
gain.inv_nodes
```

### Calculating gain based on node_caps

#### Looking at distribution in terms of three subsets by node_caps
```{r subsetting_node_caps}
subsets.node_caps <- table(breast_cancer$node_caps, breast_cancer$class)
subsets.node_caps
```

#### Computing entropy of node_caps subsets
```{r entropy_node_caps}
entropy_ncno <-entropy.gini(subsets.node_caps["no",], type=2)$entropy
entropy_ncyes <-entropy.gini(subsets.node_caps["yes",], type=2)$entropy
```

#### Computing weighted entropy for node_caps
```{r weighted_entropy_node_caps}
w_ncno <- sum(subsets.node_caps["no",])/total
w_ncyes <- sum(subsets.node_caps["yes",])/total
info.node_caps <- w_ncno*entropy_ncno + w_ncyes*entropy_ncyes
```

#### Computing information gain for node_caps
```{r information_gain_node_caps}
gain.node_caps <- info.before - info.node_caps
gain.node_caps
```

### Calculating gain based on deg_malig

#### Looking at distribution in terms of three subsets by deg_malig
```{r subsetting_deg_malig}
subsets.deg_malig <- table(breast_cancer$deg_malig, breast_cancer$class)
subsets.deg_malig
```

#### Computing entropy of deg_malig subsets
```{r entropy_deg_malig}
entropy_dm1 <-entropy.gini(subsets.deg_malig["1",], type=2)$entropy
entropy_dm2 <-entropy.gini(subsets.deg_malig["2",], type=2)$entropy
entropy_dm3 <-entropy.gini(subsets.deg_malig["3",], type=2)$entropy

```

#### Computing weighted entropy for deg_malig
```{r weighted_entropy_deg_malig}
w_dm1 <- sum(subsets.deg_malig["1",])/total
w_dm2 <- sum(subsets.deg_malig["2",])/total
w_dm3 <- sum(subsets.deg_malig["3",])/total
info.deg_malig <- w_dm1*entropy_dm1 + w_dm2*entropy_dm2 + w_dm3*entropy_dm3
```

#### Computing information gain for deg_malig
```{r information_gain_deg_malig}
gain.deg_malig <- info.before - info.deg_malig
gain.deg_malig
```

### Calculating gain based on breast

#### Looking at distribution in terms of three subsets by breast
```{r subsetting_breast}
subsets.breast <- table(breast_cancer$breast, breast_cancer$class)
subsets.breast
```

#### Computing entropy of breast subsets
```{r entropy_breast}
entropy_left <-entropy.gini(subsets.breast["left",], type=2)$entropy
entropy_right <-entropy.gini(subsets.breast["right",], type=2)$entropy
```

#### Computing weighted entropy for breast
```{r weighted_entropy_breast}
w_left <- sum(subsets.breast["left",])/total
w_right <- sum(subsets.breast["right",])/total
info.breast <- w_left*entropy_left + w_right*entropy_right
```

#### Computing information gain for breast
```{r information_gain_breast}
gain.breast <- info.before - info.breast
gain.breast
```

### Calculating gain based on breast quadrant

#### Looking at distribution in terms of three subsets by breast_quad
```{r subsetting_breast_quad}
subsets.breast_quad <- table(breast_cancer$breast_quad, breast_cancer$class)
subsets.breast_quad
```

#### Computing entropy of breast quadrant subsets
```{r entropy_breast_quad}
entropy_bq01 <-entropy.gini(subsets.breast_quad["central",], type=2)$entropy
entropy_bq02 <-entropy.gini(subsets.breast_quad["left_low",], type=2)$entropy
entropy_bq03 <-entropy.gini(subsets.breast_quad["left_up",], type=2)$entropy
entropy_bq04 <-entropy.gini(subsets.breast_quad["right_low",], type=2)$entropy
entropy_bq05 <-entropy.gini(subsets.breast_quad["right_up",], type=2)$entropy
```

#### Computing weighted entropy for breast quadrant
```{r weighted_entropy_breast_quad}
w_bq01 <- sum(subsets.breast_quad["central",])/total
w_bq02 <- sum(subsets.breast_quad["left_low",])/total
w_bq03 <- sum(subsets.breast_quad["left_up",])/total
w_bq04 <- sum(subsets.breast_quad["right_low",])/total
w_bq05 <- sum(subsets.breast_quad["right_up",])/total
info.breast_quad <- w_bq01*entropy_bq01 + w_bq02*entropy_bq02 + w_bq03*entropy_bq03 + w_bq04*entropy_bq04 + w_bq05*entropy_bq05
```

#### Computing information gain for breast quadrant
```{r information_gain_breast_quad}
gain.breast_quad <- info.before - info.breast_quad
gain.breast_quad
```

### Calculating gain based on irradiation

#### Looking at distribution in terms of three subsets by irradiation
```{r subsetting_irradiat}
subsets.irradiat <- table(breast_cancer$irradiat, breast_cancer$class)
subsets.irradiat
```

#### Computing entropy of irradiation subsets
```{r entropy_irradiat}
entropy_irrno <-entropy.gini(subsets.irradiat["no",], type=2)$entropy
entropy_irryes <-entropy.gini(subsets.irradiat["yes",], type=2)$entropy
```

#### Computing weighted entropy for irradiation
```{r weighted_entropy_irradiat}
w_irrno <- sum(subsets.irradiat["no",])/total
w_irryes <- sum(subsets.irradiat["yes",])/total
info.irradiat <- w_irrno*entropy_irrno + w_irryes*entropy_irryes
```

#### Computing information gain for irradiation
```{r information_gain_irradiat}
gain.irradiat <- info.before - info.irradiat
gain.irradiat
```

### Results: Choosing root node
```{r choose_root_node}
# Creating vector of gains for evaluation
gains <- c(gain.age, gain.breast, gain.breast_quad, gain.deg_malig, gain.inv_nodes, gain.irradiat, gain.menopause, gain.node_caps, gain.tumor_size)

# Creating vector of names
gain_names <- c("age", "breast", "breast_quad", "deg_malig", "inv_nodes", "irradiat", "menopause", "node_caps", "tumor_size")

# Creating named vector of gains
names(gains) <- gain_names

# Calculating maximum information gain
# The attribute with the maximum information gain, which will serve as the root node in our decision tree, is:
names(gains)[which.max(gains)]
```

