---
title: "INFO-I 421 Final - Building Decision Tree Models for the Breast Cancer Data"
author: "Nicholas Bibeau"
date: "April 29, 2019"
output: 
  html_document:
    toc: TRUE
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rpart)
library(rpart.plot)
library(DiagrammeR)
library(useful)
library(xgboost)
library(vip)
library(randomForest)
```

# 1.1 Importing Data
```{r data_import}
breast_cancer <- as_tibble(read.table("https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer/breast-cancer.data", header = FALSE, na = "?", sep = ",", col.names=c("class", "age", "menopause", "tumor_size", "inv_nodes", "node_caps", "deg_malig", "breast", "breast-quad", "irradiat")))
```

# 1.2 Finding the age category which has the highest risk of recurrence

## Viewing distinct factors for age
```{r unique_age}
unique(breast_cancer$age)
```

## Building contingency table for age and recurrence
```{r age_recur_table}
breast_cancer_table <- table(breast_cancer$age, breast_cancer$class, dnn = c("Age", "Recurrence"))
breast_cancer_table
```

## Calculating probability of recurrence by age group

```{r prob_recur_by_age}
probAge1 <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(age == "20-29")) /
  count(breast_cancer %>% filter(age == "20-29"))
probAge2 <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(age == "30-39")) /
  count(breast_cancer %>% filter(age == "30-39"))
probAge3 <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(age == "40-49")) /
  count(breast_cancer %>% filter(age == "40-49"))
probAge4 <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(age == "50-59")) /
  count(breast_cancer %>% filter(age == "50-59"))
probAge5 <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(age == "60-69")) /
  count(breast_cancer %>% filter(age == "60-69"))
probAge6 <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(age == "70-79")) /
  count(breast_cancer %>% filter(age == "70-79"))

# Creating vector of recurrence probability by age group
recur_probs <- c(probAge1, probAge2, probAge3, probAge4, probAge5, probAge6)

# Creating vector of distinct age groups
ages <- c ("20-29", "30-39", "40-49", "50-59", "60-69", "70-79")

# Creating named vector of recurrence probability by age group
names(recur_probs) <- ages

```

## Determining age category with the highest risk of recurrence
```{r max_risk_age} 
# The age group with the highest risk of recurrence is: 
names(recur_probs)[which.max(recur_probs)]
```

# 1.3 Finding breast with the higher risk of recurrence

## Viewing distinct factors for breast
```{r unique_breast}
unique(breast_cancer$breast)
```

## Building contingency table for age and recurrence
```{r breast_recur_table}
breast_cancer_table <- table(breast_cancer$breast, breast_cancer$class, dnn = c("Breast", "Recurrence"))
breast_cancer_table
```

## Calculating probability of recurrence by breast

```{r prob_recur_by_breast}
probLeft <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(breast == "left")) /
  count(breast_cancer %>% filter(breast == "left"))
probRight <- count(breast_cancer %>% filter(class == "recurrence-events") %>% filter(breast == "right")) /
  count(breast_cancer %>% filter(breast == "right"))

# Creating vector of recurrence probability by breast
recur_probs_breasts <- c(probLeft, probRight)

# Creating vector of distinct breasts
breasts <- c ("left", "right")

# Creating named vector of recurrence probability by age group
names(recur_probs_breasts) <- breasts

```

## Determining breast with the higher risk of recurrence
```{r max_risk_breast} 
# The age group with the highest risk of recurrence is: 
names(recur_probs_breasts)[which.max(recur_probs_breasts)]
```

# 1.4 Drop missing values
```{r drop_missing}
breast_cancer <- breast_cancer %>% drop_na()
```

# 1.5 Rank attributes and plot correlation matrix
