---
title: "Predictive Learning with Linear Regression"
author: "Nicholas Bibeau"
date: "March 20, 2019"
output: 
  html_document:
    toc: TRUE
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(polycor)
library(corrplot)
```


# Load Dataset
```{r}
#Load the cpu dataset
library("tidyverse")
cpu <- as_tibble(read_csv("https://www.iun.edu/~cisjw/ds/files/data/cpu.csv"))
#change attribute names to lowercase
colnames(cpu) <- tolower(colnames(cpu))
summary(cpu)

```

# Ready data for analysis

## Split the set into two sets for the training and testing purposes

```{r}
set.seed(52)
rows <- sample(nrow(cpu)) #Shuffle row indices
rows

cpu <- cpu[rows, ] #Randomly reorder the cpu rows by the vector rows
cpu
```

## Subset data
```{r}
split <- round(nrow(cpu)*0.66) #138

# Put first 66% into train dataset
train <- cpu[1:split, ]

# Put remainining 1/3 into test dataset
test <- cpu[(split+1):nrow(cpu), ]
```

# Fit model
```{r}
model <- lm(performance ~ ., train) 

summary(model)
```

## Extracting coefficients and r-squared
```{r}
#Extract Coefficients
model$coefficients #Or coef(model)

#Read r-squared
summary(model)$r.squared
```

# Plot fitted values vs. true values in the training set
```{r}
plot(model$fitted.values, train$performance, xlab="true_performance", ylab="fitted_performance", main="t")
```

#  Calculate RMSE in the training set
```{r}
# Calculating the errors between every true value and its fitted value in the training set and store the errors in a vector error:
error <- model$fitted.values - train$performance

# Calculating RMSE:
rmse.train <- sqrt(mean(error^2))
rmse.train
```

# Applying the model to the test set
```{r}
# We can apply the model to the test set test by the predict function. Store the predictions in a vector called scores:
scores <- predict(model, test)
scores
```

# Calculating Error Metric: RMSE (root mean square error)
```{r}
# Calculating the errors between every true value and its fitted value in the training set and store the errors in a vector error:
error <- scores - test$performance
# Calculating RMSE:
rmse.test <- sqrt(mean(error^2))
rmse.test
```

# Comparing rmse.train and rmse.test
```{r}
# The training set has a lower RMSE because the regression model was created using this data set.
```

