---
title: "Week 8 - Correlation Analysis"
author: "Nicholas Bibeau"
date: "February 25, 2019"
output: 
  html_document:
    toc: TRUE
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# 12.4 Exercise

## Importing weather.csv as a tibble
```{r}
weather <- read_csv("https://www.iun.edu/~cisjw/ds/files/data/weather.csv")
```

## Creating normalization function
```{r}
normalize <- function(v, range=1){
    v.norm = (v - min(na.omit(v))) / (max(na.omit(v))-min(na.omit(v))) * range
    return(v.norm)
}

normalizeDataset <- function (data){
  data.norm <- data
  types <- sapply(data, is.numeric)
  for (i in 1:length(types)) {
    if(types[i]==TRUE){
      data.norm[ ,i] <- normalize(data[ ,i])
    }
  }
  return(data.norm)
}
```


## Normalizing numeric column data in weather tibble
```{r}
weather.norm <- normalizeDataset(weather)
weather.norm
```


## Creating function to evaluate Pearson correlation
```{r}
pairCorr <- function (x,y) {
  # calculate the Pearson correlation between two columns in X and Y  
  length <- length(x)
  mu_x <- mean(x)
  mu_y <- mean(y)
  sd_x <- sd(x)
  sd_y <- sd(y)
  # calculate the numerator
  sum <- 0
  for (i in 1:length) {
    sum <- sum + (x[i]-mu_x)*(y[i]-mu_y)
  }
  corr <- sum/((length-1)*sd_x*sd_y)
  return(corr)
}
```

## Computing Pearson correlation for temperature and humidity
```{r}
#The Pearson correlation between temperature and humidity is
pairCorr(weather.norm$temperature, weather.norm$humidity)
```

## Results
```{r}
# A Pearson correlation of 0.3150818 indicates no correlation between our variables, temperature and humidity.
```

# 12.6 Exercise

## Importing train.csv as a tibble
```{r}
train <- read_csv("https://www.iun.edu/~cisjw/ds/files/data/train.csv")
```

## Inspecting values for Survived and Sex in the train tibble
```{r}
# Viewing disctinct values in Survived column
train %>% distinct(Survived)

# Viewing distinct values in Sex column
train %>% distinct(Sex)
```

## Creating contingency table
```{r}
# Creating contingency table for Survived and Sex
train_table <- table(train$Survived, train$Sex, dnn = c("Survived", "Sex"))

#Viewing contingency table
train_table

# Viewing chi-square test
summary(train_table)
```

## Conclusion
```{r}
# The χ2 value for 1 degree of freedom required to reject the null hypothesis
```




Find whether there is a correlation relationship between Survived and Sex in the Titanic dataset, train.csv at https://www.iun.edu/~cisjw/ds/files/data/.

Write R code to print the contingendy table between the attributes Survived and Sex, perform a Chi-square test between the two attributes.

Make a conclusion on the correlation between Survived and Sex, according to the Chi-square value and p-value.