---
title: "Week 6 Data manipulation with tibble and dplyr"
author: "Nicholas Bibeau"
date: "February 17, 2019"
output: 
  html_document:
    toc: TRUE
    theme: united
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# 8.1 E1 - Use dplyr and pipes. Rewrite the following scripts:

## mtcars <- data.frame(mtcars)
```{r}
mtcars <- mtcars %>% as_tibble
```

## mtcars[1:5, ]
```{r}
mtcars %>% slice(1:5)
```

## tail(mtcars, 5)
```{r}
mtcars %>% slice(28:32)
```


## nrow(mtcars)
```{r}
mtcars %>% count()
```

## mtcars[ , c("mpg")]
```{r}
mtcars %>% select(mpg)
```

## mtcars[mtcars$cyl == "6","mpg"]
```{r}
mtcars %>% filter(cyl == 6) %>% select(mpg)
```

## mtcars[mtcars$cyl == "6", ]
```{r}
mtcars %>% filter(cyl == 6)
```

## mtcars[mtcars$mpg > "25", c("mpg","cyl")]
```{r}
mtcars %>% filter(mpg > 25) %>% select(c("mpg", "cyl"))
```

# 8.2 E2 - Use dplyr and pipes. Rewrite the following script:
```{r}
library(tidyverse)
diamonds <- data.frame(diamonds)
diamonds[1:5, ]
nrow(diamonds)
nrow(diamonds[diamonds$cut == "Very Good", ]) 
nrow(diamonds[diamonds$carat > "3.0", ]) 
diamonds[diamonds$color == "D", c("color","cut")]
mean(diamonds$price)
```

```{r}
library(tidyverse)
diamonds <- diamonds %>% data.frame
diamonds %>% slice(1:5)
diamonds %>% count()
diamonds %>% filter(cut == "Very Good") %>% count()
diamonds %>% filter(carat > 3.0) %>% count()
diamonds %>% filter(color == "D") %>% select(c("color", "cut"))
diamonds %>% summarize(mean(price))

```

# 8.3 E3

## 1. Write code to group the mtcars data by cyl and find the total of each group.
```{r}
mtcars %>% group_by(cyl) %>% summarize(n())
```

## 2. Write code to group the mtcars data by cyl and find mean mpg and mean disp of each group.
```{r}
mtcars %>% group_by(cyl) %>% summarize(mean(mpg), mean(disp))
```

## 3. Write code to group the diamonds data by cut and find the maximum and minimum price for each group.
```{r}
diamonds %>% group_by(cut) %>% summarize(max(price), min(price))
```

## 4. Write code to group the diamonds data by color and find the maximum, mean, minimum price for each group.
```{r}
diamonds %>% group_by(color) %>% summarize(max(price), mean(price), min(price))
```